The brightness and spatial distributions of terrestrial radio sources

### 2013-02-28 articles/Article-RFI-distribution/article-rfi-distribution.tex
Acknowledgments and final touch

G.~Macario$^{21}$, G.~Mann$^{17}$, J.~P.~McKean$^{4}$, H.~Meulman$^{4}$, M.~J.~Norden$^{4}$, 

@@

\begin{abstract}
Faint undetected sources of radio-frequency interference (RFI) might become visible after long time integration of radio observations if they are stationary and consistently present over time. Thereby, they might obstruct the detection of the weak astronomical signals of interest. This issue is especially important for Epoch of Reionisation (EoR) projects that try to detect the faint redshifted HI signals from the time of the earliest structures in the Universe. We explore the low-frequency RFI situation by studying brightness histograms of visibility data observed with LOFAR, similar to radio-source-count analyses that are used in cosmology. The histograms show an RFI distribution that follows a power-law distribution with an estimated exponent around -1.5. With several assumptions, this can be explained with a uniform distribution of terrestrial radio sources whose radiation follows existing propagation models. Extrapolation of the power law would imply that the current LOFAR EoR observations should be severely RFI limited if these sources add up consistently. These observations almost reach the thermal noise and are thought not to be limited by RFI. Therefore, we conclude that it is likely that most RFI sources do not linearly increase the noise, in which case RFI will not prevent an EoR detection with LOFAR.
\end{abstract}

@@

\section{Introduction}
Radio astronomy concerns itself with the observation of radiation from celestial sources at radio wavelengths. However, astronomical radio observations can be affected by radio-frequency interference (RFI), which makes it difficult to calibrate the instrument and achieve high sensitivities \citep{impact-of-warc79,interference-and-radioastronomy-1991,interference-model-lemmon,rfi-mitigation-overview-fridman-baan}. Many techniques have been designed to mitigate this effect, such as detection and flagging of data \citep{chi-square-time-blanking-weber, multichannel-rfi-mitigation, exoplanet-detection-with-rfi, wsrt-rfims, pulse-blanking, effelsberg-rfi-mitigation, LOFAR-RFI-pipeline}, adaptive cancellation techniques \citep{adaptive-cancellation,post-correlation-reference-signal} and spatial filtering \citep{multichannel-rfi-mitigation, ellingson-spatial-nulling-2002, hampson-spatial-nulling-2002, boonstra-dissertation, spatial-filtering-parkes-multibeam-for-pulses, post-correlation-filtering}. In most cases, and with the help of careful spectrum allocation management and radio-quiet zones, these methods excise enough of the interference to allow calibration and imaging of the data.

@@

For the EoR project with $1$~MHz resolution, Eq.~\ref{eq:rfi-noise-level} results in $\sigma_\textrm{RFI}\approx 4.7$ mJy. However, the first EoR results of observations of one day have approximately reached the thermal noise of about $1.7$ mJy per 0.2~MHz sub-band (Yatawatta et al. 2013, \textit{in press}), and the resulting images show no signs of RFI. This implies that either the upper limit is far from the actual RFI situation, or Eq.~\ref{eq:rfi-noise-level} is not applicable to most of the RFI that is observed with LOFAR. In the following section we will discuss effects that could cause a reduced contribution of RFI.

@@

Because of the above two arguments, when considering RFI it is a risk to use the NCP as one of the EoR target fields. At the same time, this field is useful for analysing the RFI coherency properties. It is also a simple field to observe with LOFAR, because it is always at high declination in the Netherlands and it does not contain bright foreground sources. Preliminary analysis of EoR NCP observations of a single night have reached the thermal noise (Yatawatta et al. 2013, \textit{in press}), but do not show leaked RFI at the pole.

@@

\section*{Acknowledgments}
LOFAR, the Low-Frequency Array designed and constructed by ASTRON, has facilities in several countries, that are owned by various parties (each with their own funding sources), and that are collectively operated by the International LOFAR Telescope (ILT) foundation under a joint scientific policy.

>>>>

G.~Maccario$^{21}$, G.~Mann$^{17}$, J.~P.~McKean$^{4}$, H.~Meulman$^{4}$, M.~J.~Norden$^{4}$, 

@@

\begin{abstract}
Faint undetected sources of radio-frequency interference (RFI) might become visible after long time integration of radio observations if they are stationary and consistently present over time. Thereby, they might obstruct the detection of the weak astronomical signals of interest. This issue is especially important for Epoch of Reionisation (EoR) projects that try to detect the faint redshifted HI signals from the time of the earliest structures in the Universe. We explore the low-frequency RFI situation by studying brightness histograms of visibility data observed with LOFAR, similar to radio-source-count analyses that are used in cosmology. The histograms show an RFI distribution that follows a power-law distribution with an estimated exponent around -1.5. With several assumptions, this can be explained with a uniform distribution of terrestrial radio sources whose radiation follows existing propagation models. Extrapolation of the power law would imply that the current LOFAR EoR observations should be severely RFI limited if these sources add up consistently. This is in contrast with the actual observations that almost reach the thermal noise and are thought not to be limited by RFI. Therefore, we conclude that it is likely that most RFI sources do not linearly increase the noise, in which case RFI will not prevent an EoR detection with LOFAR.
\end{abstract}

@@

\section{Introduction}
Radio astronomy concerns itself with the observation of radiation from celestial sources at radio wavelengths. However, astronomical radio observations can be affected by radio-frequency interference (RFI), which makes it difficult to calibrate the instrument and achieve high sensitivities \citep{impact-of-warc79,interference-and-radioastronomy-1991,interference-model-lemmon,rfi-mitigation-overview-fridman-baan}. Many techniques have been designed to mitigate this effect, such as detection and flagging of data \citep{chi-square-time-blanking-weber, multichannel-rfi-mitigation, exoplanet-detection-with-rfi, wsrt-rfims, pulse-blanking, effelsberg-rfi-mitigation, post-correlation-rfi-classification}, adaptive cancellation techniques \citep{adaptive-cancellation,post-correlation-reference-signal} and spatial filtering \citep{multichannel-rfi-mitigation, ellingson-spatial-nulling-2002, hampson-spatial-nulling-2002, boonstra-dissertation, spatial-filtering-parkes-multibeam-for-pulses, post-correlation-filtering}. In most cases, and with the help of careful spectrum allocation management and radio-quiet zones, these methods excise enough of the interference to allow calibration and imaging of the data.

@@

For the EoR project with $1$~MHz resolution, Eq.~\ref{eq:rfi-noise-level} results in $\sigma_\textrm{RFI}\approx 4.7$ mJy. However, the first EoR results of observations of one day have approximately reached the thermal noise of about $1.7$ mJy per 0.2~MHz sub-band \citep{ncp-eor-yatawatta}, and the resulting images show no signs of RFI. This implies that either the upper limit is far from the actual RFI situation, or Eq.~\ref{eq:rfi-noise-level} is not applicable to most of the RFI that is observed with LOFAR. In the following section we will discuss effects that could cause a reduced contribution of RFI.

@@

Because of the above two arguments, when considering RFI it is a risk to use the NCP as one of the EoR target fields. At the same time, this field is useful for analysing the RFI coherency properties. It is also a simple field to observe with LOFAR, because it is always at high declination in the Netherlands and it does not contain bright foreground sources. Preliminary analysis of EoR NCP observations of a single night have reached the thermal noise \citep{ncp-eor-yatawatta}, but do not show leaked RFI at the pole.

@@

\section*{Acknowledgments}
LOFAR, the Low-Frequency Array designed and constructed by ASTRON, has facilities in several countries, that are owned by various parties (each with their own funding sources), and that are collectively operated by the International LOFAR Telescope (ILT) foundation under a joint scientific policy. Parts of this research were conducted by the Australian Research Council Centre of Excellence for All-sky Astrophysics (CAASTRO), through project number CE110001020. C. Ferrari and G. Maccario acknowledge financial support by the ``Agence Nationale de la Recherche'' through grant ANR-09-JCJC-0001-01.
